---
title: "codebook"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


**STEP 1**
Download required files to working directory and unzip them. 
[get files](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones)

**STEP 2**
Get path to working directory
```{r,  message=FALSE, warning=FALSE}
path <- file.path(getwd(), "UCI HAR Dataset")

```


**STEP 3**

Include the required Libraries
```{r,  message=FALSE, warning=FALSE}

library(data.table)
library(dplyr)
library(tidyr)
library(reshape2)
library(knitr)

```
**STEP 4**
Read train subject and test subject file into R. 
```{r,message=FALSE, warning=FALSE}
train_subject <- read.table(file.path(path, "train", "subject_train.txt"))
test_subject <- read.table(file.path(path, "test", "subject_test.txt"))

```
Both Subject are merged
```{r,message=FALSE, warning=FALSE}
merge_subject <- bind_rows(train_subject, test_subject)
```
Column renamed to aubject
```{r}
names(merge_subject) <- "subject"
```
**STEP 5**

Step 4 repeated for Labels(activity)

```{r}
train_label <- read.table(file.path(path, "train", "Y_train.txt"))
test_label <- read.table(file.path(path, "test", "y_test.txt"))
merge_label <- bind_rows(train_label, test_label)
names(merge_label) <- "activity" 

```
**STEP 6**

Step 4 repeated for set( data)

```{r}
train_set <- read.table(file.path(path, "train", "x_train.txt"))
test_set <- read.table(file.path(path, "test", "x_test.txt"))
merge_set <- bind_rows(train_set, test_set)

```
**STEP 7**
All column are merged
```{r}
merge_data <- bind_cols(merge_subject, merge_label, merge_set)


```

**STEP 8**
Data frame is converted to data table and then sorted

```{r}
merge_data <- as.data.table(merge_data)
setkey(merge_data, subject, activity)
```

```{r, echo=FALSE}
kable(head(merge_data), caption = "Sample of sorted merged data" )
```
**STEP 9**

Melting is done to change wide-format data  into long-format data

```{r}
merge_data <- data.table(melt(merge_data, key(merge_data), variable.name="variable"))


```
```{r, echo=FALSE}
kable(head(merge_data), caption = "Sample of melted merged data", align = "c")
```


**STEP 10**
Extract only the measurements on the mean and standard deviation 
for each measurement
"features.text" file which contains List of all features read into R. Column are renamed.The value of variable is changed to match variable of merge_data so merging can be done
```{r}
features <- read.table(file.path(path, "features.txt"))
names(features) <- c("variable", "features") # Rename column
features$variable<- paste0("V", features$variable)# adjust value of variable
merge_data2 <- merge(merge_data, features, by = "variable") #merge both data

```


**STEP 11**
Subset mean and standar deviation(std) from merged data

```{r}
subset_data <- merge_data2[grepl("subject|label|mean\\(\\)|std\\(\\)", merge_data2$features),]


```

**STEP 12**
Use descriptive activity names to name the activities in the data set
"activity_labels.txt" file which contains decription for label is read loaded into R and columns are renamed
```{r}
activity_label <- read.table(file.path(path, "activity_labels.txt"))
names(activity_label) <- c("activity", "activitiName")## rename column

```

**STEP 13**
The subset data table is merged with the activity label
```{r}
subset_data_description<- merge(subset_data, activity_label, by = "activity")


```
`````{r, echo=FALSE}
kable(head(subset_data_description), caption = "Merged data with feature column added and aktivity name added", align = "c")

```

**STEP 14**
Appropriately label the data set with descriptive variable names.
a function "f" is used here to separate features into multiple variables
```{r}
f<- function (A) {
  grepl(A, subset_data_description$features)
}
```

Get domain, instrument, acceleration, mean and std 
```{r}
 y <- matrix(1:2)
x <- matrix(c(f("^t"), f("^f")), ncol=nrow(y))
subset_data_description$domain <- factor(x %*% y, labels=c("Time", "Freq"))


x <- matrix(c(f("Acc"), f("Gyro")), ncol=nrow(y))
subset_data_description$instrument <- factor(x %*% y, labels=c("accelerometer", "gyroscope"))

x <- matrix(c(f("BodyAcc"), f("GravityAcc")), ncol=nrow(y))
subset_data_description$acceleration <- factor(x %*% y, labels=c(NA, "body", "gravity"))

x <- matrix(c(f("mean()"), f("std()")), ncol=nrow(y))
subset_data_description$variable <- factor(x %*% y, labels=c("mean", "std"))


```
Get Jerk and magnitude
```{r}
subset_data_description$jerk <- factor(f("Jerk"), labels=c(NA, "jerk"))
subset_data_description$magnitude <- factor(f("Mag"), labels=c(NA, "magnitude"))

```
Get XYZ axis
```{r}
y <- matrix(1:3, nrow=3)
x <- matrix(c(f("-X"), f("-Y"), f("-Z")), ncol=nrow(y))
subset_data_description$axis <- factor(x %*% y, labels=c(NA, "x", "y", "z"))

```

```{r, echo=FALSE}
kable(head(subset_data_description), caption = "Data with features separated into multiple columns", align = "c")

```

**STEP 15**

Independent tidy data set with the average of each variable for each activity and each subject is created. collumns are rearrange and redundant columns are removed
```{r}
tidydata <-
  subset_data_description %>% 
  group_by_(.dots=c("subject", "activitiName", "instrument", "acceleration", "magnitude", "jerk", "variable")) %>% 
  summarize(average=mean(value))  
```
```{r, echo=FALSE}
kable(head(tidydata), caption = "Tidy data", align = "c")
```



